---
title: "Peer Assessment 1"
author: "Pedro Carneiro Junior"
date: "Saturday, September 19, 2015"
output: html_document
---

* * *

## Introduction

The main objective of this document is to fulfill the requirements of the Coursera's Peer Assessment 1 of the Reproducible Research course (part of the Data Science Specialization track), which can be found at:

- [https://class.coursera.org/repdata-032/human_grading/view/courses/975145/assessments/3/submissions], for students currently enrolled at the Reproducible Research course and the respective teachers.

* * *

## Resolution


#### Setting global options


**knitr and opts_chunk**

In our specific case, it is recommended to set global configurations as from what is asked in the requirements.

For that, we need to set echo=TRUE globaly so that anybody will be able to read all the code chunks after processing the results. But, for having a working opts_chunk() function it is necessary that the knitr package is loaded inside the R Markdown first.

```{r setoptions, echo=TRUE}
library("knitr")
opts_chunk$set(echo=TRUE)
```


#### Loading necessary packages


**sqldf**

We are going to use the sqldf package because of our preference to SQL syntax for data manipulation. So, first we test if the sqldf package is installed and if not, run the command to install it. Then we load the sqldf package to memory.

```{r InstallAndLoadSQLDF}
# If needed, installing, then loading necessary packages...
packages <- c("sqldf")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}
library("sqldf")
```

### Loading and preprocessing the data

I decided to download the data from the source https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip, anyhow a copy of the original remains in my forked Github repository at https://github.com/pedrokarneiro/RepData_PeerAssessment1.

#### Loading data

As the data source file is in CSV format, then it is better to use the read.csv() function.

First, let's inform the user about where his/her data is going to be stored and processed.

```{r ShowWorkingDirectory}
print(paste("A T E N T I O N :"
            , "This is your current working directory and where all project files will be located: "
            , getwd()))
```

Afterwards, it is healthy to clean memory in order to avoid low memory problems.

```{r ClearMemory}
# Clearing memory for calculation safety...
rm(list=ls())
```

Then we get the file from the given URL and unzip it to the working directory. In fact, as https protocol does not fit directly into every knitr environment, specially under Windows 7 and RStudio, R produces "unsupported URL scheme" error while trying to download a zip file, it is wise to follow the recommendations informed at [http://stackoverflow.com/questions/25341285/error-when-knitr-has-to-download-a-zip-file].

Note that changing the URL to http and setting file.download to mode="wb" the problem is resolved and the script can be knitted successfully.

```{r DownloadingZipFile}
fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destFilePath <- paste(getwd(), "/repdata%2Fdata%2Factivity.zip", sep="")
download.file(fileURL, destfile=destFilePath, mode="wb")
unzip("repdata%2Fdata%2Factivity.zip")
list.files()
#dateDownloaded <- date()
#dateDownloaded
```

Finally, data is loaded into memory.

```{r LoadingData}
activityCSV <- read.csv("activity.csv")
```

#### First Analysis of the Data

In order to process/transform the data (if necessary) into a format suitable for analysis, let's explore some information about the dataset.

Notice that there are many NAs for the steps variable (2304 in total out of 17568 observations), and that indicates us that they may affect results and that we will have to deal with them.

```{r ExploreDataHeadAndTail}
head(activityCSV)
tail(activityCSV)
```

```{r ExploreDataSummary}
summary(activityCSV)
```

```{r ExploreDataStructure}
str(activityCSV)
```

* * *

### What is mean total number of steps taken per day?

As recommended, we are going to ignore the NA values for the following tasks:

- Calculate the total number of steps taken per day
- Calculate and report the mean and median of the total number of steps taken per day
- Make a histogram of the total number of steps taken each day

**Calculating the total number of steps taken per day**

To calculate the mean we are going to need the number of total steps per day. So we need to select the sum of steps SUM(steps) from the activityCSV data frame where the steps are not NA and then group the sums by date.

```{r TotalStepsPerDay}
totalStepsPerDay <- sqldf("SELECT date, SUM(steps) AS sum_steps_day
                           FROM activityCSV
                           WHERE steps <> 'NA'
                           GROUP BY date")
print(totalStepsPerDay)
```

**Calculating and report the mean and median of the total number of steps taken per day**

```{r CalculateMeanAndMedian}
myMean <- mean(totalStepsPerDay$sum_steps_day)
myMedian <- median(totalStepsPerDay$sum_steps_day)
print(paste("The daily steps calculated mean is ", myMean, " and median is ", myMedian, sep=""))
```

**Making a histogram of the total number of steps taken each day**

To define number of breaks (how many bar will appear in the histogram) we need to count how many observations our totalStepsPerDay has.

```{r DefiningBreaks}
myBreaks <- sqldf("SELECT COUNT(*) AS days
                   FROM totalStepsPerDay")
print(myBreaks)
```

Now, plotting the histogram:

```{r DailyStepsHistogram}
myMain <- "Total Number of Steps Taken Each Day"
myYlab <- "Frequency"
myBreaks <- myBreaks$days
myXlab <- "Total Steps per Day"

hist(totalStepsPerDay$sum_steps_day
     , main = myMain
     , ylab = myYlab
     , xlab = myXlab
     , breaks = myBreaks
     , col = rainbow(myBreaks)
)
abline(v=myMedian, col="red")
mtext(text = paste("The red line near ", round(myMedian)," indicates the median", sep=""), side = 3, col = "red")
```

### What is the average daily activity pattern?

Now, there are two tasks:

- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

**Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days**

```{r AvgStepsPerInterval}
meanStepsPerInterval <- aggregate(steps ~ interval, activityCSV, mean)
plot(meanStepsPerInterval
     , type = "l"
     , main = ("Daily Average of Steps per Interval")
     , xlab = "5-minute intervals"
     , ylab = "average number of steps taken, averaged across all days"
     , col="blue")
```

**Reporting the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps**

```{r RptMaxSteps}
maxAvgStepsInterval <- which.max(meanStepsPerInterval$steps)
maxStepsInterval <- meanStepsPerInterval[maxAvgStepsInterval,]$interval
maxSteps <- meanStepsPerInterval[maxAvgStepsInterval,]$steps

# Plotting again just to show in the report
plot(meanStepsPerInterval
     , type = "l"
     , main = ("Daily Average of Steps per Interval")
     , xlab = "5-minute intervals"
     , ylab = "average number of steps taken, averaged across all days"
     , col="blue")

abline(v=maxStepsInterval, col="red")
mtext(text = paste("The red line on ", maxStepsInterval
                   , " indicates the maximum number of steps (~"
                   , round(maxSteps, digits = 2), ")"
                   , sep="")
      , side = 3, col = "red")

print(paste("The maximum number of steps is, on average (rounded to two digits precision), ", round(maxSteps, digits = 2), ", and it happens on the 5-minute interval number ", maxStepsInterval, sep=""))

```

### Imputing missing values

There are four tasks here:

- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

- Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

- Create a new dataset that is equal to the original dataset but with the missing data filled in.

- Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


***Calculating and reporting the total number of missing values in the dataset (i.e. the total number of rows with NAs)***
```{r CalculateNAs}
incompleteCases <- activityCSV[!complete.cases(activityCSV),]
numberOfIncompleteCases <- nrow(incompleteCases)
print(paste("The number of incomplete cases (NAs) is: ", numberOfIncompleteCases, sep=""))
```

***Devising a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.***

***Creating a new dataset that is equal to the original dataset but with the missing data filled in.***

In order to fill the NAs with approximate values, we decided to change the sequence of the execution for the benefit of what is asked: First we create a new data frame named **withoutNAs** just as a copy of **activityCSV**, then we proceed replacing each NA by the mean of steps in the interval, which, in our case is represented by the data frame **meanStepsPerInterval**. For each row of **withoutNAs** (a copy of the original data frame still with NAs) and for each *steps* that is NA, we get the corresponding *interval* and then get the *steps* value from **meanStepsPerInterval** (the data frame with the steps averages) for the same *interval*, replacing the *steps* NA value with it.


```{r FillNAs}
# perform the imputation

withoutNAs <- activityCSV

for (i in 1:nrow(withoutNAs)){
  if (is.na(withoutNAs$steps[i])){
    activityInterval <- withoutNAs$interval[i]
    intervalMatch <- which(meanStepsPerInterval$interval == activityInterval)
    intlAvgSteps <- meanStepsPerInterval$steps[intervalMatch]
    withoutNAs$steps[i] <- intlAvgSteps
  }
}

print("Bellow we provide a small comparison between the two datasets: the original activityCSV (with NAs) and withoutNAs (the same data frame, but withoutNAs)")
head(activityCSV)
head(withoutNAs)
```

***Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?***

```{r TotalStepsPerDay2}
totalStepsPerDay2 <- sqldf("SELECT date, SUM(steps) AS sum_steps_day
                           FROM withoutNAs
                           GROUP BY date")
print(totalStepsPerDay2)
```

Let's calculate and report the mean and median of the total number of steps taken per day of withouNAs to compare with the data from activityCSV.

```{r CalculateMeanAndMedian2}
myMean2 <- mean(totalStepsPerDay$sum_steps_day)
myMedian2 <- median(totalStepsPerDay$sum_steps_day)
print(paste("For activityCSV, the daily steps calculated mean is ", myMean, " and median is ", myMedian, sep=""))
print(paste("For withoutNAs, the daily steps calculated mean is ", myMean2, " and median is ", myMedian2, sep=""))
```

**Making a histogram of the total number of steps taken each day for the new data frame without NAs (withoutNAs)**

To define number of breaks (how many bar will appear in the histogram) we need to count how many observations our totalStepsPerDay2 has.

```{r DefiningBreaks2}
myBreaks2 <- sqldf("SELECT COUNT(*) AS days
                   FROM totalStepsPerDay2")
print(myBreaks2)
```

Now, plotting the histogram for withoutNAs and re-plotting the histogram for the original activityCSV:

```{r DailyStepsHistogram2}
myMain2 <- "Total Number of Steps Taken Each Day (without NAs)"
myYlab2 <- "Frequency"
myBreaks2 <- myBreaks2$days
myXlab2 <- "Total Steps per Day"

hist(totalStepsPerDay2$sum_steps_day
     , main = myMain2
     , ylab = myYlab2
     , xlab = myXlab2
     , breaks = myBreaks2
     , col = "blue"
)
abline(v=myMedian2, col="red")
mtext(text = paste("The red line near ", round(myMedian2)," indicates the median", sep=""), side = 3, col = "red")

hist(totalStepsPerDay$sum_steps_day
     , main = myMain
     , ylab = myYlab
     , xlab = myXlab
     , breaks = myBreaks
     , col = rainbow(myBreaks)
)
abline(v=myMedian, col="red")
mtext(text = paste("The red line near ", round(myMedian)," indicates the median", sep=""), side = 3, col = "red")

```

- Do these values differ from the estimates from the first part of the assignment?

Yes, despite that the mean and the median do not change, the values from the new dataset without NAs differ from the original dataset with NAs, as observed by comparing the two plots, specially regarding to the position of the greatest frequency in relation to the median.

- What is the impact of imputing missing data on the estimates of the total daily number of steps?

The main impact, in my evaluation, is that missing data on the estimates will give a false perspective of the frequency and of the greatest frequency in relation to the median.

**So it is advised that one should not include NAs on the estimates of the total daily number of steps, in other words, the blue histogram is better (more trustworthy) than the rainbow histogram.**

### Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

UNDER CONSTRUCTION

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

UNDER CONSTRUCTION

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

UNDER CONSTRUCTION

